Combining information from photometric and spectroscopic data:\\ Probabilistic spectrophotometric inference

### 2015-07-12 tex/ms.tex
start reorganizing for consistent runs and mocks, switching to linear fitting, fix strings for some plots
changes to mock generation script for ease of use, consistency

The Right Thing To Do is to express the \emph{reason} that the
spectroscopic data are less reliable \emph{within} the
likelihood function.
That is, the one must add nuisance parameters to parameterize the
unreliability of the spectroscopy, and infer those along with the
parameters of interest.
If this parameterization is good, the information in the spectroscopy
data will not flow entirely to the parameters of interest; instead,
some will flow to the nuisance parameters.
This will reduce the importance of the spectroscopy relative to the
photometry.
This amounts to propogating the uncertainties in the
spectrophotometric calibration (or alternatively in the continuum
normalization).
In what follows, we provide an example of this kind of
parameterization and inference.

@@

If you don't believe the reported data noise variances $\sigma_n^2$,
it often makes sense to add in quadrature a ``jitter'' variance $s^2$.
This is a simple modification of the chi-squared log-likelihood:
\begin{eqnarray}\label{eq:photometricLF}
\ln p(d\given\theta,s^2) &=& -\frac{1}{2}\,\sum_{n=1}^N
                             \left[\frac{[d_n - \mu_n(\theta)]^2}
                             {[\sigma_n^2 + s^2]} + 
                             \ln(2\pi\,[\sigma_n^2 + s^2]) \right]
\quad .
\end{eqnarray}
This modification shows the necessity of including the log-variance
term.  Large $s$ reduces the penalty for individual large residuals,
but also reduces the reward for well-fit data.
This---equation~(\ref{eq:photometricLF})---is in fact the
log-likelihood function we are going to use for the photometric
measurements in what follows.  For photometric data this ``jitter'' is
often best described as a fractional error, $s^2 \equiv
\hat{s}^2\mu(\theta)^2$, corresponding for example to an additional
magnitude zero-point uncertainty.  In this formulation, if $\theta$
are the parameters of interest, the jitter variance $\hat{s} =
s/\mu(\theta)$ is a nuisance noise-model parameter representing
additonal fractional uncertainty in the flux.

{\color{blue} BDJ This paragraph should explain why we are using the
GP, not just how it works.}
For the spectroscopic measurements we are going to account for the
spectrophotometric calibration in our likelihood function.  We take
the calibration-vector that modifies the mean spectroscopic model to
produce the observed spectrum to be multiplicative with respect to the
mean model.  The multiplicative calibration vector can be explicitly
modeled, for example using polynomials.  However, for complex
calibration vectors over large wavelength ranges, the number of
parameters required to model the calibration can become quite large
(e.g. very high order polynomials) and the correct parameterization
may be unknown.  It is therefore desirable to introduce a
`non-parameteric' model of the calibration.  We do this using a
Gaussian Process in the likelihood function.  The Gaussian Process is
a generalization of the modified log-likelihood introduced in
Equation \label{eq:photometricLF} that allows for correlated noise in
the residuals, as would be introduced by inaccurate calibration.

@@

In detail, we are going to choose the exponential of a $M$th-order
polynomial form for the calibration vector $f(\alpha)$ and a bell
shape (Gaussian) for the kernel function $K_\alpha(\cdot)$:
\begin{eqnarray}\displaystyle
f(\alpha) &=& \exp\left(\sum_{m=o}^M c_m\,x^m\right)
\\
x &\equiv& \frac{\lambda}{\lambda_0} - 1
\\
K_\alpha(\lambda_n - \lambda_{n'}) &=& \mu_n\,\mu_{n'}\, a^2\,\exp\left(-\frac{[\lambda_n - \lambda_{n'}]^2}{2\,\ell^2}\right)
\\
\transpose{\alpha} &\equiv& \left[ \{c_m\}_{m=0}^M, s^2, a^2, \ell^2 \right]
\quad ,
\end{eqnarray}
where $\lambda$ is a column vector made up of the wavelengths
$\lambda_m$ and $x$ is a scaled and shifted version of the same,
scaled by a fiducial wavelength $\lambda_0$.  The exponentiation of
the polynomial ensures that $f(\alpha)$ is everywhere positive.
Nuisance parameters $c_m$ set the shape of the large-scale calibration
issues, while $a^2$ and $\ell^2$ set the variance and correlation
length of the remaining covariant calibration noise. Structure in the
residuals on the scale of $\ell$ and with amplitude $a$ will not be
strongly penalized, but the determinant of $C$ becomes larger for
larger $a$ and smaller $\ell$. The multiplication by $\mu_n\,
\mu_{n'}$ in the kernel function signifies that it is defined in
\emph{fractional} terms.  Using a high-order polynomial can result in
strong degeneracies between model and polynomial parameters, whereas
the GP {\color{blue} BDJ make this more precise and correct} only
reduces the likelihood penalty for residuals on the scale of $\ell$
with amplitude $a$ that {\it can't} be fit by the mean model.

@@

Next we will \emph{not} assume that the calibration is perfectly known
and we will therefore allow the calibration parameters $\alpha$ in our
model to vary.  {\color{red} This is not converged.} In Figure
\ref{fig:speconly_calibrated} we show the result of our inference from
the spectrum and a single photometric measurement (SDSS $g$) for a
perfectly calibrated mock spectrum. To demonstrate the quality of the
fit we show both the mock input spectrum and predictions of the
observed spectra that result from several draws of the model
parameters $[\theta, \alpha]$ from posterior distribution.  We also
show samples of the inferred calibration vector,
\begin{eqnarray} \label{eq:calibration}
\frac{d}{\mu(\theta)} & = & f(\alpha) + \frac{\tilde{\Delta}}{\mu(\theta)}
\quad ,
\end{eqnarray}
where $\tilde{\Delta}$ is the Gaussian Process mean prediction for the
residual between the data and the model \citep{RW},
defined here as 
\begin{eqnarray} \label{eq:calibration}
\tilde{\Delta} & = & K \, C^{-1} \, \Delta
\end{eqnarray}.  
For the perfectly calibrated spectrum the true calibration vector
(shown in black) is one everywhere.  The spread of the inferred
calibration vectors approximately indicates our uncertainty in the
spectrophotometric calibration when no additional calibration
information is available and only a single photometric point is
available.

When the spectrophotometric calibration is assumed to be unknown,
the posterior PDFs for stellar mass and especially $A_v$ become
extremely large.  There is now no information about the overall shape
of the SED (and hence the dust attenuation), and the overall
normalization is constrained entirely by the single photometric point.
The uncertainty in the stellar mass is somewhat broader than the
uncertainty on the $g$-band flux because of the variation in
mass-to-light ratio due to uncertainties in dust and, to a lesser
extent, age.

@@

In Figure \ref{fig:speconly_uncalibrated} we show the results of
inference on an \emph{uncalibrated} mock spectrum.  The results are
similar to the case of the \emph{calibrated} spectrum. An important
difference is that the inferred calibration vector now contains numerous small scale wiggles,
similar to the actual calibration vector.  The overall shape of the
spectrum is still unconstrained, leading to large degeneracies between
the polynomial terms and the dust attenuation.

>>>>

The Right Thing To Do is to express the \emph{reason} that the
spectroscopic data are less reliable \emph{within} the
likelihood function.
That is, the one must add nuisance parameters to parameterize the
unreliability of the spectroscopy, and infer those along with the
parameters of interest.
If this parameterization is good, the information in the spectroscopy
data will not flow entirely to the parameters of interest; instead,
some will flow to the nuisance parameters.
This will reduce the importance of the spectroscopy relative to the
photometry.
This amounts to propagating the uncertainties in the
spectrophotometric calibration (or alternatively in the continuum
normalization).
In what follows, we provide an example of this kind of
parameterization and inference.

@@

If you don't believe the reported data noise variances $\sigma_n^2$,
it often makes sense to add in quadrature a ``jitter'' variance $b^2$.
This is a simple modification of the chi-squared log-likelihood:
\begin{eqnarray}\label{eq:photometricLF}
\ln p(d\given\theta,b^2) &=& -\frac{1}{2}\,\sum_{n=1}^N
                             \left[\frac{[d_n - \mu_n(\theta)]^2}
                             {[\sigma_n^2 + b^2]} + 
                             \ln(2\pi\,[\sigma_n^2 + b^2]) \right]
\quad .
\end{eqnarray}
This modification shows the necessity of including the log-variance
term.  Large $b$ reduces the penalty for individual large residuals,
but also reduces the reward for well-fit data.
This---equation~(\ref{eq:photometricLF})---is in fact the
log-likelihood function we are going to use for the photometric
measurements in what follows.  For photometric data this ``jitter'' is
often best described as a fractional error, $b^2 \equiv
\hat{b}^2\mu(\theta)^2$, corresponding for example to an additional
magnitude zero-point uncertainty.  In this formulation, if $\theta$
are the parameters of interest, the jitter variance $\hat{b} =
b/\mu(\theta)$ is a nuisance noise-model parameter representing
additonal fractional uncertainty in the flux.

{\color{blue} BDJ This paragraph should explain why we are using the
GP, not just how it works.}
For the spectroscopic measurements we are going to account for the
spectrophotometric calibration in our likelihood function.  We take
the calibration-vector that modifies the mean spectroscopic model to
produce the observed spectrum to be multiplicative with respect to the
mean model.  The multiplicative calibration vector can be explicitly
modeled, for example using polynomials.  However, for complex
calibration vectors over large wavelength ranges, the number of
parameters required to model the calibration can become quite large
(e.g. very high order polynomials) and the correct parameterization
may be unknown.  It is therefore desirable to introduce a
`non-parameteric' model of the calibration.  We do this using a
Gaussian Process in the likelihood function.  The Gaussian Process is
a generalization of the modified log-likelihood introduced in
equation~\label{eq:photometricLF} that allows for correlated noise in
the residuals, as would be introduced by inaccurate calibration.

@@

In detail, we are going to choose the exponential of a $M$th-order
polynomial form for the calibration vector $f(\alpha)$ and a bell
shape (Gaussian) for the kernel function $K_\alpha(\cdot)$:
\begin{eqnarray}\displaystyle
f(\alpha) &=& \exp\left(\sum_{m=o}^M c_m\,x^m\right)
\\
x &\equiv& \frac{\lambda}{\lambda_0} - 1
\\
K_\alpha(\lambda_n - \lambda_{n'}) &=& \mu_n\,\mu_{n'}\, a^2\,\exp\left(-\frac{[\lambda_n - \lambda_{n'}]^2}{2\,\ell^2}\right)
\\
\transpose{\alpha} &\equiv& \left[ \{c_m\}_{m=0}^M, s^2, a^2, \ell^2 \right]
\quad ,
\end{eqnarray}
where $\lambda$ is a column vector made up of the wavelengths
$\lambda_m$ and $x$ is a scaled and shifted version of the same,
scaled by a fiducial wavelength $\lambda_0$.  The exponentiation of
the polynomial ensures that $f(\alpha)$ is everywhere positive.
Nuisance parameters $c_m$ set the shape of the large-scale calibration
issues, while $a^2$ and $\ell^2$ set the variance and correlation
length of the remaining covariant calibration noise. Structure in the
residuals on the scale of $\ell$ and with amplitude $a$ will not be
strongly penalized, but the determinant of $C$ becomes larger for
larger $a$ and smaller $\ell$. The multiplication by $\mu_n\,
\mu_{n'}$ in the kernel function signifies that it is defined in
\emph{fractional} terms.  Using a high-order polynomial can result in
strong degeneracies between model and polynomial parameters, whereas
the Gaussian Process {\color{blue} BDJ make this more precise and
correct} likelihood only reduces the penalty for residuals on the
scale of $\ell$ with amplitude $a$ that {\it can't} be fit by the mean
model.

The two log-likelihoods, for photometry and spectroscopy, are combined
into a single likelihood function,
\begin{eqnarray}\displaystyle ln p(d_s, d_p\given \theta, \alpha, b) =
ln p_{spec}(d_s\given \theta, alpha) + ln p_{phot}(d_p\given \theta,
b) \quad .
\end{eqnarray} where now $d_s$ is the spectroscopic data and $d_p$ is
the photometric data.  The spectroscipc likelihood $p_{spec}$ is given
by equation~(\ref{eq:spectroscopicLF}) and $p_{phot}$ is given by
equation~(\ref{eq:photometricLF}).

@@

Next we will \emph{not} assume that the calibration is perfectly known
and we will therefore allow the calibration parameters $\alpha$ in our
model to vary.  {\color{red} This is not converged.} In Figure
\ref{fig:speconly_calibrated} we show the result of our inference from
the spectrum and a single photometric measurement (SDSS $g$) for a
perfectly calibrated mock spectrum. To demonstrate the quality of the
fit we show both the mock input spectrum and predictions of the
observed spectra that result from several draws of the model
parameters $[\theta, \alpha]$ from posterior distribution.  We also
show samples of the inferred calibration vector,
\begin{eqnarray} \label{eq:calibration}
\frac{d}{\mu(\theta)} & = & f(\alpha) + \frac{\tilde{\Delta}}{\mu(\theta)}
\quad ,
\end{eqnarray}
where $\tilde{\Delta}$ is the Gaussian Process mean prediction for the
residual between the data and the model \citep{RW},
defined here as 
\begin{eqnarray} \label{eq:calibration}
\tilde{\Delta} & = & K \, C^{-1} \, \Delta
\quad .
\end{eqnarray}
For the perfectly calibrated spectrum the true calibration vector
(shown in black) is one everywhere.  The spread of the inferred
calibration vectors approximately indicates our uncertainty in the
spectrophotometric calibration when no additional calibration
information is available and only a single photometric point is
available.

When the spectrophotometric calibration is assumed to be unknown and
is explicitly modeled, as in Figure \ref{fig:speconly_calibrated}, the
posterior PDFs for stellar mass and especially $A_v$ become extremely
large.  There is now no information about the overall shape of the SED
(and hence the dust attenuation), and the overall normalization is
constrained entirely by the single photometric point.  The fractional
uncertainty in the stellar mass is somewhat broader than the
fractional uncertainty on the $g$-band flux because of the variation
in mass-to-light ratio due to uncertainties in dust and, to a lesser
extent, age.

@@

In Figure \ref{fig:speconly_uncalibrated} we show the results of
inference on an \emph{uncalibrated} mock spectrum.  The results are
similar to the case of the \emph{calibrated} spectrum. An important
difference is that the inferred calibration vector now contains
numerous small scale wiggles, similar to the actual calibration vector
used to generate the mock.  The overall shape of the spectrum is still
unconstrained, leading to large degeneracies between the polynomial
terms and the dust attenuation, and a poor constraint on the dust
attenuation and stellar mass.
